---
output:
  md_document:
    variant: markdown_github
---

[![Build Status](https://travis-ci.org/raquamaps/aquamapsdata.svg?branch=master)](https://travis-ci.org/raquamaps/aquamapsdata)  

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r include=FALSE}
knitr::opts_chunk$set(warning=FALSE, comment=NA)
```

```console
Welcome to ...
                           ▌   ▐
▝▀▖▞▀▌▌ ▌▝▀▖▛▚▀▖▝▀▖▛▀▖▞▀▘▞▀▌▝▀▖▜▀ ▝▀▖
▞▀▌▚▄▌▌ ▌▞▀▌▌▐ ▌▞▀▌▙▄▘▝▀▖▌ ▌▞▀▌▐ ▖▞▀▌
▝▀▘  ▌▝▀▘▝▀▘▘▝ ▘▝▀▘▌  ▀▀ ▝▀▘▝▀▘ ▀ ▝▀▘
```

## Introduction

`aquamapsdata` is an R package that can download and create a local SQLite database with datasets from AquaMaps.org (2017)

These datasets are available to web browsers through https://aquamaps.org, but the `aquamapsdata` package offers an a couple of convenient functions for accessing this data in an IDE like `RStudio`.

## Installing from github

If you want to install the latest version of the `aquamapsdata` package from github, you can do it like so:

```{r message=FALSE, warning=FALSE, results="hide", eval=FALSE}
# First make sure you have the devtools package
# which simplifies installations from github
# Note: Windows users have to first install Rtools to use devtools

install.packages("devtools") 

library("devtools")
install_git("https://github.com/raquamaps/aquamapsdata.git", build_vignettes=FALSE)

```

## Quick start

Load the package in your R environment:

```{r message=FALSE, warning=FALSE, results="hide", eval=FALSE}

library(aquamapsdata)

download_db()

occ()
nativemaps()
taxa()
hcaf()
hspen()

```

To see some quick usage examples to get you started, open the Vignette.

The package uses SQLite3, which means you need to install it.


```console
# on linux
sudo apt install sqlite3

# on mac
brew install sqlite3
```

## Credits

`aquamapsdata` would have not been possible without many amazing R libraries, such as

* "tidyverse" ie dplyr, tidyr, stringr etc from Hadley Wickham
* rgbif etc from ROpenSci

## Meta

* Please [report any issues or bugs](https://github.com/raquamaps/aquamapsdata/issues).
* License: AGPL
